Using an iterated block particle filter via \pkg{spatPomp}

### 2023-06-08 ms.Rnw
tidying new diagnostics section

If likelihood maximization appears to be failing, or has unacceptably high Monte Carlo variability, one wishes to diagnose the cause, or causes.
Even if inference appears to be operating satisfactorily, it is good practice to see whether there are some data points for which the model is inappropriate, or some parameters for which optimization is less stable.
It may be interesting to know if less Monte Carlo effort would be sufficient, or if more would be useful.
For all these questions, a starting point is exploratory analysis of diagnostic data produced by \code{ibpf} and \code{bpfilter}.

The success of an iterated filter for likelihood maximization depends on the success of a single pass of that filter for likelihood evaluation.
A useful diagnostic quantity for the particle filter is the effective sample size (ESS) \citep{douce11,liu01}.
In the notation of \citep{king16}, the filter observation weight for particle $j$ at time $n$ is $w(n,j)=f_{Y_n|X_n}(y_n^*|X^P_{n,j};\theta)$, the normal weight is $\tilde{w}(n,j)=w(n,j)\big/ \sum_{k=1}^Jw(n,k)$, and the effective sample size is
\begin{equation}
\mathrm{ESS}_n = \left(\sum_{j=1}^J \tilde{w}(n,j)^2\right)^{-1}.
\end{equation}
ESS is generally motivated as an approximation to the equivalent number of independent samples from the filtering distribution. 
In the context of likelihood-based inference, it is convenient to think of $\mathrm{ESS}^{-1}_n$ as an approximation to the variance of the conditional log-likelihood estimate,
\begin{equation}
\hat \ell_n = \log \left( \frac{1}{J} \sum_{j=1}^J w(n,j) \right).
\end{equation}
However, ubiquitous multi-core computation permits us to calculate this variance directly, by Monte Carlo replication.
This bypasses consideration of when ESS is, and is not, a good estimator. 
Also, the $R$ Monte Carlo replicates, $\hat\ell_n^{(r)}$, $r=1,\dots,R$, give rise to an improved estimate of the conditional log-likelihood,
\begin{equation}
\bar \ell_n = \frac{1}{R} \sum_{r=1}^R \hat\ell_n^{(r)},
\end{equation}
together with an uncertainty estimate via the central limit theorem.

@@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observation.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution, and so we prefer to work with CLLV.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citep{asfaw23}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@

@@

>>>>

If likelihood maximization appears to be failing, or has unacceptably high Monte Carlo variability, one wishes to diagnose the cause, or causes.
Even if inference appears to be operating satisfactorily, it is good practice to see whether there are some data points for which the model is inappropriate, or some parameters for which optimization is less stable.
It may be interesting to know if less Monte Carlo effort would be sufficient, or if greater effort would be beneficial.
For all these questions, a starting point is exploratory analysis of diagnostic data produced by \code{ibpf} and \code{bpfilter}.

The success of an iterated filter for likelihood maximization depends on the success of a single pass of that filter for likelihood evaluation.
A useful diagnostic quantity for the particle filter is the effective sample size (ESS) \citep{doucet11,liu01}.
In the notation of \citet{king16}, the filter observation weight for particle $j$ at time $n$ is $w(n,j)=f_{Y_n|X_n}(y_n^*|X^P_{n,j};\theta)$, the normal weight is $\tilde{w}(n,j)=w(n,j)\big/ \sum_{k=1}^Jw(n,k)$, and the effective sample size is
\begin{equation}
\mathrm{ESS}_n = \left(\sum_{j=1}^J \tilde{w}(n,j)^2\right)^{-1}.
\end{equation}
ESS is generally motivated as an approximation to the equivalent number of independent samples from the filtering distribution. 
In the context of likelihood-based inference, it is convenient to think of $\mathrm{ESS}^{-1}_n$ as an approximation to the variance of the conditional log-likelihood estimate,
\begin{equation}
\hat \ell_n = \log \left( \frac{1}{J} \sum_{j=1}^J w(n,j) \right).
\end{equation}
However, ubiquitous multi-core computation permits us to calculate this variance directly, by Monte Carlo replication.
This bypasses consideration of whether or not ESS is a good estimator. 
Also, the Monte Carlo replicates, denoted by $\hat\ell_n^{(r)}$, for $r=1,\dots,R$, give rise to an improved estimate of the conditional log-likelihood,
\begin{equation}
\label{eq:ell-bar}
\bar \ell_n = \frac{1}{R} \sum_{r=1}^R \hat\ell_n^{(r)},
\end{equation}
together with an uncertainty estimate via the central limit theorem.
One could also replace the arithmetic average  in (\ref{eq:ell-bar}) using \code{pomp::logmeanexp}.

@@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observation.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citet{asfaw23arxiv}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@

@@

Log-likelihood anomalies can be plotted against any covariate of interest for exploratory data analysis.
This is analogous to residual analysis for linear models.
